---
title: "Data Visualization Activity 2"
author: "Dr. Luis Malpica"
date: "March, 2020"
output:
  pdf_document:
    toc: yes
  html_document:
    theme: united
    toc: yes
editor_options: 
  chunk_output_type: console
---

# Introduction

Now it is your turn to make a publication quality figure, your goal is to polish a plot for publication. 

Get the two plots below ready for publication, ideally in the same multipanel plot, remember:

- Check: Axes, data density, theme, overplotting, etc.
- Output publication quality
- Make a figure caption

# Install required packages
```{r, message=FALSE}

library(tidyverse)
library(lubridate)
library(ggthemes)
library(gridExtra)

```

# Load, check & clean data
```{r}
perch <- read.csv("data/YPerchTL.csv")

# To make things faster just run these lines:
perch$sampledate <- mdy(perch$sampledate)

perch <- perch %>%
  select(-lakeid) %>%
  select(-spname) %>% 
  select(-year4) %>% 
  mutate(year = year(sampledate))

# We just assigned as data 'sampledate' and removed unnecessary columns
# STEP 1, did it load correctly?
head(perch)

# STEP 2: Are the data types right?
sapply(perch, class)

# STEP 3: Check for missing or impossible values
range(perch$sampledate)

range(perch$year)

range(perch$length)
plot(perch$length)
hist(perch$length)

range(perch$weight)

# let's remove those nasty NAs that are messing up our data
range(perch$weight, na.rm = T)
plot(perch$weight)
hist(perch$weight)

# STEP 4: Factor levels
levels(perch$gearid)

# Seems allright
```

# PLOT 1 with ggplot
```{r}
# This plot should present the weight over time, perhaps sample date such as this one:
perchplot <- ggplot(perch, aes(x=sampledate, y=weight)) + 
  geom_point()

perchplot
```

# PLOT 2 with ggplot
```{r}
# This plot should present the weight by gear type, such as this one:
perchplot2 <- ggplot(perch, aes(x=gearid, y=weight)) +
  geom_boxplot()

perchplot2
```

# Saving image
Try to save plots on a single image.
```{r}
ggsave("./plots/Figure3.pdf", perchplot, 
       dpi=300, 
       width = 16, height = 18,
       device = "pdf",
       units = "cm",
       useDingbats=FALSE) 

ggsave("./plots/Figure4.pdf", perchplot2, 
       dpi=300, 
       width = 16, height = 18,
       device = "pdf",
       units = "cm",
       useDingbats=FALSE) 


```

